---
title: "Homework 04: Tidy Data and Joins"
author: "Aidan Hughes"
date: "October 9, 2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Overview

We were tasked to choose one prompt for reshaping data and one for joining. Here are the two that I chose:

1. *"Make a tibble with one row per year and columns for life expectancy for two or more countries."*

2. *"Create a second data frame, complementary to Gapminder. Join this with (part of) Gapminder using a dplyr join function and make some observations about the process and result. Explore the different types of joins."*

As always, import packages first.

```{r packages}
suppressPackageStartupMessages(library("gapminder"))
suppressPackageStartupMessages(library("tidyverse"))
```


# Reshaping: *"Make a tibble with one row per year and columns for life expectancy for two or more countries."*

Let's make a new tibble from the Gapminder dataframe with a subset of a few random countries. We'll print the data as a table in long (or "tidy") format first.

```{r tidy table}
lifeExpTidy <- gapminder %>%
  select(year, country, lifeExp) %>%
  filter(country %in% c("Argentina", "Canada", "Germany", "Japan", "Mexico", "Spain"))

# The tibble contains data for all years, but only print data from 1952 and 2002 just to keep the table short
lifeExpTidy %>%
  filter(year %in% c("1952", "2002")) %>%
  knitr::kable()
```

Now we can use the `spread()` function to reshape our data into wide (or "untidy") format.

```{r untidy table}
lifeExpUntidy <- lifeExpTidy %>%
  spread(key="country", value="lifeExp")
  
lifeExpUntidy %>%
  knitr::kable()
```

Using the `spread()` function makes tables more human-friendly for reading.

Now we can compare the life expectancies over time of a couple different countries with a scatterplot.

```{r lifeExp scatterplot}
lifeExpUntidy %>%
  ggplot(aes(x = year)) +
  geom_point(aes(y = Canada, color = "Canada")) +
  geom_line(aes(y = Canada, color = "Canada")) +
  geom_point(aes(y = Mexico, color = "Mexico")) +
  geom_line(aes(y = Mexico, color = "Mexico")) +
  labs(title = "Life Expectancy Vs. Time", x = "Year", y = "Life Expectancy", color="Country") +
  
  # Center the title, left-aligned by default
  theme(plot.title = element_text(hjust = 0.5))

```


# Task 2: *"Create a second data frame, complementary to Gapminder. Join this with (part of) Gapminder using a dplyr join function and make some observations about the process and result. Explore the different types of joins."*
 
Let's create a tibble from Gapminder of all countries and the continent they're in, then create a second data frame containing only the capital city of some random countries.

```{r countries and continents}
countries <- gapminder %>%
  distinct(country, .keep_all = TRUE) %>%
  select(country, continent)

countries %>%
  head(15) %>%
  knitr::kable()
```

```{r capital cities}
capitals <- countries %>%
  select(country) %>%
  filter(country %in% c("Argentina", "Canada", "Germany", "Japan", "Mexico", "Spain")) %>%
  cbind("capital" = c("Buenos Aires", "Ottawa", "Berlin", "Tokyo", "Mexico City", "Madrid"))

capitals %>%
  knitr::kable()
```

Now let's try using the different join functions on our two datasets and compare the results.

## Inner-join

Using inner-join, the countries that had capitals now also have been joined with their continents.

```{r innerjoin}
inner_join(x = countries, y = capitals) %>%
  knitr::kable()
```

## Semi-join

Using semi-join, we're left with only the countries that have capitals but don't actually join their values for `capital`.

```{r semijoin}
semi_join(x = countries, y = capitals) %>%
  knitr::kable()
```

## Left-join

Using left-join, all countries with continents are kept and are joined with their capital if they have one, otherwise capital is NA. We would expect to get the same result in a full-join as well.

```{r leftjoin}
left_join(x = countries, y = capitals) %>%
  head(15) %>%
  knitr::kable()
```

## Full-join

```{r fulljoin}
full_join(x = countries, y = capitals) %>%
  head(15) %>%
  knitr::kable()
```

## Anti-join

With anti-join, we get very different results depending on which dataset is `x` and which is `y`. If `x = countries`, we get a set of all countries without specified capitals. However, with `x = capitals`, we get no results, as there are no countries with a specified capital but no specified continent.

```{r antijoin}
anti_join(x = countries, y = capitals) %>%
  head(15) %>%
  knitr::kable()
```

```{r antijoin opposite}
anti_join(x = capitals, y = countries) %>%
  nrow()
```

